Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-30 paper2.qmd
section:6; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Among these, Case V has become the most widely used model in empirical CJ research, primarily due to the widespread adoption of the BTL model. The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008]. Although the substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021] and simplifies the measurement model, the other two core assumptions of the model can still lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023], especially when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].

Furthermore, because Thurstone primarily aimed to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory did not specify how to use the trait estimates for statistical inference. The CJ tradition has attempted to fill this gap by separating trait estimation from hypothesis testing. Specifically, researchers often rely on point estimates of the traits, such as the BTL scores or their transformations, for inference. Although this approach simplifies analysis, the statistical literature cautions that it can introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

The study then translates the SCM into a bespoke statistical model that addresses key limitations of Case V. This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2). In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli and incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical). These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

Five research avenues--current and new--deserve particular attention. The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them. Notably, while addressing these questions through formal derivations--following the tradition of classical statistical models like linear regression--may seem a natural step, we argue that the complexity of the CJ system makes this approach initially impractical. Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without relying on cumbersome mathematical proofs. Nonetheless, developing formal mathematical proofs remains an essential goal for future work.

@@

<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
Sampling and comparison mechanisms play a central role in modern CJ experimental design. However, most of the CJ literature has examined these mechanisms within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. Although these studies offer valuable insights, we argue that they may have overlooked the broader role these mechanisms play within the CJ system. Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system. To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.

@@

<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data. This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities. While some studies report improvements in trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b],  others argue that these gains are artificially inflated by the method [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

<!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework. This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design. Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022]. To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons--one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Building on this and the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such characteristics indeed undermine the "sample-freeness" assumption, then it becomes essential to explore strategies for controlling their influence. This includes investigating how many judges--and how many judgments per judge--are required to produce reliable trait estimates, as well as whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge in this general formulation: the measurement model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Among these, Case V remains the most widely used model in empirical CJ research, mainly due to the widespread adoption of the BTL model. The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008]. Notably, although these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Moreover, because Thurstone's original goal was to produce a "coarse scaling" of  traits and allocate stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory offered no guidance on how to use trait estimates for statistical inference. To address this gap, researchers in the CJ tradition have often separated trait estimation from hypothesis testing, relying on point estimates of the traits--such as BTL scores or their transformations--for inference. Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

The study then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2). In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical). By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

Five research avenues, both current and new, deserve particular attention. The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them. Notably, although it may seem like a natural next step to address these questions through formal derivations, following the tradition of classical statistical models like linear regression, we contend that the complexity of the CJ system makes this approach impractical from the outset. Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without depending on cumbersome mathematical proofs. Nonetheless, developing formal mathematical proofs remains an essential goal for future work.

@@

<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
Although sampling and comparison mechanisms are central to modern CJ experimental design, it is striking that most of the CJ literature has examined them only within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. Although these studies offer valuable insights, we argue that they have overlooked the broader role these mechanisms play within the CJ system. Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system. To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.

@@

<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data. This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities. While some studies report that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

<!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions, an approach supported by our framework. This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design. Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@Morin_et_al_2018] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022]. To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
Moreover, if we accept that judges may prioritize different aspects of a stimulus, it follows that their assessments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Thus, building on the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such characteristics indeed undermine the assumption of "sample-freeness," then it becomes essential to explore strategies to control their influence and to determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts. Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
